Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-08-04 manuscript/manuscript.tex
edits to text, especially Introduction

Low-frequency gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \sv{citation}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
The dominant astrophysical source in this frequency range is an isotropic stochastic background 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}.

Most measuring of the stochastic background is done through Bayesian data analysis. \ki{baye baye baye stuff} There are many advantages to running full Bayesian analyses on data sets, but they are slow. The bulk of the work consists of inverting non-diagonal matrices to find the strength of Hellings-Downs cross-correlations between pulsar pairs -- the "smoking gun" of a stochastic GWB.

Determining the significance of various cross-correlations between pulsar pairs is computationally expensive, requiring multi-core computing clusters many days to complete. As PTA data-analysis expands to include more parameters when measuring our predicted signal, e.g. increase in residuals with continued and new pulsar observations, uncertainty in solar system ephemerides, etc., computation time will only increase. Thus methods which hasten the production of results without compromising accuracy are not only preferable, but will be necessary. Here we explore the noise-marginalized optimal statistic as an estimator for GWB signals.

The optimal statistic is a frequentist estimator for the stochastic background amplitude. 
It was derived by \citet{abc+2009} \sv{blah blah blah... try not to plagiarize myself}

The optimal statistic is an important complement to Bayesian detection techniques. 
It provides an independent way to assess the significance of the stochastic background, 
and it is significantly faster to compute. It is also easy to test different spatial correlations 
by altering the overlap reduction function (ORF) $\Gamma$ in the cross-correlations. 
In contrast, doing a full Bayesian analysis with spatial correlations is a very computationally 
intensive procedure.

However, previous work has found that the optimal statistic gives biased results \sv{expand...} 
In order to compute the optimal statistic, the timing model parameters are fixed 
at the maximum-likelihood values, but this leads to biases in the 
optimal statistic value and uncertainty due to covariances between the 
individual red noise parameters and the GW amplitude.

In this paper, we present a technique for improving the accuracy of the optimal statistic 
by marginalizing the optimal statistic over the individual red noise parameters 
using the posterior distributions from a full Bayesian analysis. 
This hybrid approach produces a more accurate 
estimate of \Agw\ and its uncertainty.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic 
and compare to the fixed-noise optimal statistic. 
In Sec.~\ref{sec:spatial} determine how well 
the noise-marginalized optimal statistic can 
differentiate between different spatial correlations 
using simulated PTA data. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016}. 
and compare to the Bayesian analysis done by \citet{tlb+2017}. 
We conclude in Sec.~\ref{sec:conclusion}.

@@

As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
The likelihood function is given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}} &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
\sv{something about the ORF...}

@@

When constructing the optimal statistic, we typically fix the red noise parameters $\phi$ to the values that 
maximize the likelihood, which we determine from a Bayesian common red noise analysis 
\sv{check this part}. However, 
doing this biases the optimal statistic and SNR \sv{blah blah blah} 
In this paper we introduce a procedure to correct this bias by 
marginalizing over the red noise parameters using 
posterior distributions from a Bayesian analysis.

In order to compare the noise-marginalized optimal statistic to the 
fixed-noise optimal statistic, we injected a GW stochastic background with 
$\Agw = 5 \times 10^{-15}$ into simulated PTA data. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}, 
which consisted of 10 MSPs with \sv{Steve fill in here...}

The optimal statistic was computed using the \sv{fill in...} in PAL2 \sv{add citation}. 
We incorporate the individual pulsar noise parameters as follows. 
We perform individual Bayesian noise analyses on all of the pulsars, and 
fix the white noise to the maximum likelihood value. 
To compute the fixed-noise optimal statistic, 
the red noise parameters for each pulsar are set to the mode of 
the joint posterior distribution for $A_\mathrm{red}$,$\gamma_\mathrm{red}$, 
found by performing a Bayesian common red noise analysis. 
For the noise-marginalized optimal statistic, we draw from 
the common red noise posteriors in order to average the optimal statistic over 
the individual pulsars' red noise.

@@

Figure~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}

>>>>

Low-frequency gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \sv{citation}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by computing the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival.

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated 
models; however, they are computationally intensive, 
particularly when searching for evidence of the spatial correlations characteristic of a 
stochastic GWB -- ``the smoking gun'' of a stochastic GWB.

The significance of a stochastic GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude. 
It is an important complement to Bayesian detection techniques 
because it provides an independent detection procedure, 
and it is significantly faster to compute. 
In particular, different spatial correlations can be tested with the optimal statistic 
in less than a second, 
whereas a Bayesian analysis with spatial correlations requires 
several weeks to run on a supercomputing cluster.

However, previous work has found that the optimal statistic gives biased results 
for PTAs with significant red noise in the individual pulsars due to the 
strong covariance between the individual red noise parameters 
and the GWB amplitude \sv{has this ever been discussed in a paper?}. 
In this paper, we present a technique for improving the accuracy of the optimal statistic 
by marginalizing the optimal statistic over the individual red noise parameters 
using the posterior distributions from a full Bayesian analysis. 
This hybrid approach produces a more accurate 
estimate of \Agw\ and its uncertainty. 
This method is more computationally intensive than previous methods 
for computing the optimal statistic, but still requires only minutes. 
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic with any choice of the ORF, 
allowing additional spatial correlations to be tested very quickly.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic 
and compare to the fixed-noise optimal statistic. 
In Sec.~\ref{sec:spatial} determine how well 
the noise-marginalized optimal statistic can 
differentiate between different spatial correlations 
using simulated PTA data. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016}. 
We also compare our sky scramble results to the Bayesian analysis of sky scrambles done by \citet{tlb+2017}. 
We conclude in Sec.~\ref{sec:conclusion}.

@@

As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
We begin by finding the timing residuals for a given pulsar ${\bf{r}}_I$, which can be written as 
\begin{equation}
	{\bf{r}}_I = M {\bf{\epsilon}}_I + F {\bf{a}}_I + U {\bf{j}}_I + {\bf{n}}_I \,,\\
\end{equation}
where $M$ is the design matrix, 
$\epsilon_I$ is a vector of the timing model parameter uncertainties, 
${\bf{a}}_I$ are the red noise parameters describing both individual pulsars' red noise 
and common red noise in all pulsars, 
${\bf{j}}_I$ are white noise parameters describing frequency-correlated white noise, 
and ${\bf{n}}_I$ are white noise parameters describing uncorrelated white noise. 
The likelihood function is constructed from the residuals as given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}}_I &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
In the frequency domain, these can be written as
\begin{eqnarray}
	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
\end{eqnarray}
where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
two observations, 
$\mathbf{R}$ is the timing model fitting matrix, 
$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
$P_\mathrm{gw}(f)$ is the GW power spectrum, 
and $P_I(f)$ is the power spectrum for the $I$th pulsar.

@@

When constructing the residuals, 
we typically fix the red noise parameters to the values that 
maximize the likelihood. 
However, this leads to a bias in the optimal statistic and SNR because 
the individual red noise parameters and the common red noise parameters 
are highly covariant. 
In this section, we compare three different techniques for computing the 
optimal statistic for PTAs with significant red noise. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual pulsar analyses. 
Second, we fix the individual pulsars' red noise parameters to the 
values maximum-likelihood values from a common analysis 
involving all of the pulsars in our PTA. 
For the third method, we

In this paper we introduce a procedure to correct this bias by 
marginalizing over the red noise parameters using 
posterior distributions from a Bayesian analysis.

In order to compare the noise-marginalized optimal statistic to the 
fixed-noise optimal statistic, we injected a GW stochastic background with 
$\Agw = 5 \times 10^{-15}$ into simulated PTA data. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}, 
which consisted of 10 MSPs with \sv{Steve fill in here...} 
The optimal statistic was computed using the \sv{fill in...} in PAL2 \sv{add citation}. 
We incorporate the individual pulsar noise parameters as follows. 
We perform individual Bayesian noise analyses on all of the pulsars, and 
fix the white noise to the maximum likelihood value. 
For the noise-marginalized optimal statistic, we draw the red noise parameters from 
the common red noise posteriors in order to average the optimal statistic over 
the individual pulsars' red noise. 
For comparison, we also compute the fixed-noise optimal statistic in two ways -- 
first, by fixing the red noise parameters at the maximum-likelihood values from individual pulsar analyses; 
second, by fixing the red noise parameters at the maximum-likelihood values from a common analysis.

@@

In Fig.~\ref{fig:os_datasetstats} we compare the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for 425 different realizations of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=0.9\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}
